---
title: "Assignment2: Ancestry based on genotype likelihoods"
output: html_document
date: "2023-10-06"
---

__Name: Abdullah Faqih Al Mubarok__ 

__KU id: vpx267__

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(matrixStats)
```

## Estimate allele ancestral frequencies and ancestral proportions

```{r}
## Loading data 
# Minor estimated allele frequencies
est_f_data <- read.table("outK3.fopt.gz")
names(est_f_data) <- c("K1", "K2", "K3")

# Estimated ancestry
est_q_data <- read.table("outK3.qopt")
names(est_q_data) <- c("K1", "K2", "K3")

# Pop Info 
#    20 ASW -> african-american
#    20 CEU -> european 
#    20 CHB -> han chinese beijing
#    20 MXL -> mexican 
#    20 YRI -> yoruba
pop_info <- read.table("pop.info")
names(pop_info) <- c("pop_id", "indv_id")
```


To know what popuplation group represented by each column, we look at the table summary for the maximum ancestry proportion summary table  
```{r}
max_q <- tibble(max_K = colnames(est_q_data)[apply(est_q_data,1,which.max)],
                    pop_id = pop_info$pop_id)

# The 3 ancestral populations can roughly be interpreted as European, African and (Asian + Native American)  
table(max_q)
```


From the results above, we can conclude that the K1 represents African, K2 represents European,  and K3 represents (Asian + Native American). Then, we can calculate the average estimated European ancestry (K2) in the African Americans (ASW)

 - What is the average estimated European ancestry in the African Americans?
```{r}
summary_data <- est_q_data |> 
  mutate(pop_id = pop_info$pop_id, .before=1) |> 
  group_by(pop_id) |> 
  summarise(across(everything(), mean),
            .groups = 'drop')   |> 
  rename(afr_anc_prop = K1,
         eu_anc_prop = K2,
         asia_america_anc_prop = K3)

  
summary_data |> 
  filter(pop_id=="ASW")
  
```
From the data above, we can see that the average (mean) of the estimated European ancestry in the African Americans is 0.1965986. 

- For each of the five populations calculate the average estimated ancestry proportion belonging
to the 3 ancestral populations. Summaries the results in a table.
```{r}
summary_data
```


## Genotype calling based on different priors

```{r}
#calculated glik 
calc_glik <- read.table("input.gz", header = TRUE)

# true geno 
true_geno <- read.table("input.geno")

head(calc_glik)
head(true_geno)
```



```{r}
# Load input calculated Genotype Likelihoods
input_gl_ind7 <- calc_glik  |> 
  select(marker, allele1, allele2, Ind7, Ind7.1, Ind7.2)

# Indv info
info_ind7 <- pop_info |> filter(indv_id=="NA19700")

# True geno
true_geno_ind7 <- true_geno[, "NA19700"]

head(input_gl_ind7)
print(info_ind7)
table(true_geno_ind7)
```

- Assume a uniform prior (assume all 3 genotypes are equally likely). Plot a histogram of the 50000 posterior probabilities for the called genotypes.

The posterior probability for each site $j$ would be: 
$$P(G=g|X_{j}) = \frac{P(X_{j}|G=g)P(G=g)}{P(X_{j})}$$

where $P(X_{j}|G=g$ is genotype likelihood for each site $j$ for genotype $g$, $P(G=g')=\frac{1}{3}$ is the uniform prior for all genotypes $g$, and $P(X_{j}) = \sum_{g’}p(X_{ij}|g’)p(g’)$ 

Then, we call the genotype which has the maximum posterior. Here we call the first occurrence (smallest genotype velue) if there are several same values of posterior 

```{r}
# Uniform prior for 0, 1, 2 genotypes
prior <-rep(1/3,3) 

input_gl_ind7_uniform <- input_gl_ind7 |>
  # looping through each rows of dataframe
  rowwise() |> 
  mutate(posterior = (Ind7*prior[1])/(Ind7*prior[1] + Ind7.1*prior[2] + Ind7.2*prior[3]),
         posterior.1 = (Ind7.1*prior[2])/(Ind7*prior[1] + Ind7.1*prior[2] + Ind7.2*prior[3]),
         posterior.2 = (Ind7.2*prior[3])/(Ind7*prior[1] + Ind7.1*prior[2] + Ind7.2*prior[3]),
         max_posterior = max(posterior, posterior.1, posterior.2 ),
         geno_call = which.max(c_across(starts_with("posterior"))) - 1) |> 
  ungroup()
```

```{r}
ggplot(data = input_gl_ind7_uniform, mapping = aes(x = max_posterior)) +
  geom_histogram(binwidth = 0.01) + 
  labs(title = "Histogram of Called Posterior Probability", 
       subtitle = " Uniform Distribution") +
  theme_bw()
```


- Use estimated ancestral allele frequency of the African ancestry as a prior assuming Hardy- Weinberg equilibrium. Write the formula that you are using to obtain the posterior probability 3 and plot a histogram of the posterior probabilities for the called genotypes.


The posterior probability for each site $j$ would be: 
$$P(G=g|X_{j},f) = \frac{P(X_{j}|G=g)P(G=g|f)}{P(X_{j}|f)}$$

where $P(X_{j}|G=g)$ is the genotype likelihood for each site $j$ for genotype $g$. 

Assuming HWE, $P(G=g|f) = f^2$ if $G=2$, $P(G=g|f) = 2*f*(1-f)$ if $G=1$, and $P(G=g|f) = (1-f)^2$ if $G=0$.   

and $P(X_{j}) = \sum_{g’}p(X_{j}|G=g’)p(G=g’|f)$ 

```{r}
input_gl_ind7_hwe_afr <- input_gl_ind7 |>
  mutate(f = est_f_data[,"K1"]) |> 
  rowwise() |>
  mutate(posterior = ( Ind7*((1-f)^2) )/( Ind7*((1-f)^2) + Ind7.1*2*f*(1-f) + Ind7.2*(f^2) ),
         posterior.1 = ( Ind7.1*2*f*(1-f) )/( Ind7*((1-f)^2) + Ind7.1*2*f*(1-f) + Ind7.2*(f^2) ),
         posterior.2 = ( Ind7.2*(f^2) )/( Ind7*((1-f)^2) + Ind7.1*2*f*(1-f) + Ind7.2*(f^2) ),
         max_posterior = max(posterior, posterior.1, posterior.2 ),
         geno_call = which.max(c_across(starts_with("posterior"))) - 1) |>
  ungroup()
```

```{r}
ggplot(data = input_gl_ind7_hwe_afr, mapping = aes(x = max_posterior)) +
  geom_histogram(binwidth = 0.01) + 
  labs(title = "Histogram of Called Posterior Probabilities",
       subtitle = "Assume HWE and Affrican as Prior") +
  theme_bw()
```




- Try to make a prior that is better for genotypes calling by using a combinations of the 3 ancestral frequencies. Write the formulation for how you make this prior and plot a histogram of the posterior probabilities for the called genotypes.

The posterior probability for each site $j$ would be: 
$$P(G=g|X_{j},h) = \frac{P(X_{j}|G=g)P(G=g|h)}{P(X_{j}|h)}$$

where $P(X_{j}|G=g)$ is the genotype likelihood for each site $j$ for genotype $g$. 

$h_{j} = P(allele|Q, F_{j}) = q_{1}f_{1,j} + q_{2}f_{2,j} + q_{3}f_{3,j}$

Assuming HWE, $P(G=g|h) = h^2$ if $G=2$, $P(G=g|h) = 2*h*(1-h)$ if $G=1$, and $P(G=g|h) = (1-h)^2$ if $G=0$.   

and $P(X_{j}) = \sum_{g’}p(X_{j}|G=g’)p(G=g’|h)$ 

```{r} 
est_q_ind7 <- est_q_data[which(pop_info$indv_id=="NA19700"), ]

h <- est_q_ind7$K1*est_f_data$K1 + est_q_ind7$K2*est_f_data$K2 + est_q_ind7$K3*est_f_data$K3

input_gl_ind7_hwe_Q <- input_gl_ind7 |>
  mutate(f = h) |> 
  rowwise() |>
  mutate(posterior = ( Ind7*((1-f)^2) )/( Ind7*((1-f)^2) + Ind7.1*2*f*(1-f) + Ind7.2*(f^2) ),
         posterior.1 = ( Ind7.1*2*f*(1-f) )/( Ind7*((1-f)^2) + Ind7.1*2*f*(1-f) + Ind7.2*(f^2) ),
         posterior.2 = ( Ind7.2*(f^2) )/( Ind7*((1-f)^2) + Ind7.1*2*f*(1-f) + Ind7.2*(f^2) ),
         max_posterior = max(posterior, posterior.1, posterior.2 ),
         geno_call = which.max(c_across(starts_with("posterior"))) - 1) |>
  ungroup()
```

```{r}
ggplot(data = input_gl_ind7_hwe_Q, mapping = aes(x = max_posterior)) +
  geom_histogram(binwidth = 0.01) + 
  labs(title = "Histogram of Called Posterior Probabilities",
       subtitle = "Assume HWE and Using All 3 Ancestral Frequencies") +
  theme_bw()
```




- Use haplotype imputation to call the genotypes. You should run Beagle on all individuals to obtain posterior probabilities for of each possible genotype. Make a histogram of the posterior probabilities for the called genotypes for the NA19700 individual


```{bash eval=FALSE}
#beagle in the same folder
BEAGLE=./beagle.jar

java -jar $BEAGLE like=input.gz out=imputation
```

```{r}
gprobsDat<-read.table("imputation.input.gz.gprobs.gz",head=T,as.is=T)

input_gl_ind7_imputation <- gprobsDat  |> 
  select(marker, alleleA, alleleB, Ind7, Ind7.1, Ind7.2) |> 
  rowwise() |>
  mutate(max_posterior = max(Ind7, Ind7.1, Ind7.2),
         geno_call = which.max(c_across(starts_with("Ind7"))) - 1) |> 
  ungroup()
  
  
head(input_gl_ind7_imputation)
```


```{r}
ggplot(data = input_gl_ind7_imputation, mapping = aes(x = max_posterior)) +
  geom_histogram(binwidth = 0.01) + 
  labs(title = "Histogram of Called Posterior Probabilities",
       subtitle = "Beagle Imputation") +
  theme_bw()
```




```{r}
plotAccuracy<-function(x,p,to=0.95,...){
    p<-p[!is.na(x)]
    x<-x[!is.na(x)]
    ord<-order(p,decreasing=T)
    missingness <- 1-1:length(x)/length(x)
    accuracy <- 1-cumsum(!x[ord])/1:length(x)
    #ignores high missingness (where there is little data)
    ignore <- missingness > to
    lines(missingness[!ignore],accuracy[!ignore],...)
}
## make plot
plot(1,xlim=c(0,0.8),ylim=c(0.95,1),col="transparent",
    xlab="Non-call Proportion",ylab="Call Accuracy", 
    main = "Comparison of Genotype Callers")


## uniform 
plotAccuracy(input_gl_ind7_uniform$geno_call == true_geno_ind7,
             input_gl_ind7_uniform$max_posterior,
             lwd=3,col="red")
## African prior 
plotAccuracy(input_gl_ind7_hwe_afr$geno_call == true_geno_ind7,
             input_gl_ind7_hwe_afr$max_posterior,
             lwd=3,col="blue")
## 3 Ancestral priors 
plotAccuracy(input_gl_ind7_hwe_Q$geno_call == true_geno_ind7,
             input_gl_ind7_imputation$max_posterior,
             lwd=3,col="green")
## imputation 
plotAccuracy(input_gl_ind7_imputation$geno_call == true_geno_ind7,
             input_gl_ind7_imputation$max_posterior,
             lwd=3,col="hotpink")


legend("bottomright", legend=c("Uniform Prior", "African Prior", "All Ancestrals Prior", "Beagle"),
       col=c("red", "blue", "green", "hotpink"), lty=1, cex=1,
       text.font=4)
```





